---
title: "Spatial interference example"
output: html_document
---

# Spatial interference
This is a small tutorial on using linear models to adjust for spatial (or network) interference. 
First, load some packages.
```{r}
# Load necessary packages
library(sf)
library(tidyverse)
library(igraph)
library(sandwich)
library(here)
```

Next, load the data. For this tutorial, we will use Longitudinal Employer-Household Dynamics data 
to represent the network denoting interference between units.
```{r}
ca_pm25 <- read.csv("data_analysis/data/ca_bg_pm25_census_2016.csv")
ca_pm25$GEOID <- str_pad(as.character(ca_pm25$GEOID), width = 12, pad = "0")

# Work with a clean dataset
ca_pm25 <- ca_pm25 %>% drop_na()
```

Load interference graph from LEHD and process it into a data structure we can save to a file. 
Since this takes a long time, we save the data to load later without needing to run this chunk.

```{r}
od_raw <- vroom::vroom("https://lehd.ces.census.gov/data/lodes/LODES7/ca/od/ca_od_main_JT00_2016.csv.gz", show_col_types = FALSE)

ca_pm25_small <- ca_pm25 %>% select(GEOID, total_pop)
# Clean the LEHD data into an interference matrix
# that represents the proportion of the population commuting from each BG into every other BG
od_bg <- od_raw %>% mutate(w_geocode = substr(w_geocode, 1, 12),
                               h_geocode = substr(h_geocode, 1, 12)) %>%
    group_by(w_geocode, h_geocode) %>%
    # We summarize for each county
    summarize(weight = sum(S000)) %>%
    ungroup() %>% 
    right_join(ca_pm25_small, by = c("h_geocode" = "GEOID"), keep = F) %>%
    # calculate proportion of commuters by population of home geocode
    mutate(weight = weight / total_pop)

# Add filler rows for geocodes that don't commute to each other
od_bg[is.na(od_bg$weight), "weight"] <- 0
od_bg[is.na(od_bg$w_geocode), "w_geocode"] <- od_bg[is.na(od_bg$w_geocode), "h_geocode"]


write.csv(od_bg, "data_analysis/data/ca_od_bg.csv", row.names = F)
```

Next, we turn the edgelist into an adjacency matrix, and generate some sample data
to demonstrate the method. Notice how we construct "interference terms" by multiplying
the interference adjacency matrix by each column (including the intercept column of ones).
```{r}
# Turn interference edgelist into adjacency matrix
od_bg <- read.csv("data_analysis/data/ca_od_bg.csv")
od_graph = graph_from_data_frame(od_bg, directed=T)
G = as_adjacency_matrix(od_graph, attr = "weight")
diag(G) = 0

# Remove irrelevant rows and columns from G that are not in ca_pm25
inds = colnames(G) %in% substr(ca_pm25$GEOID, 2, 13)
G = G[inds, inds]

# Generate interference terms
ca_pm25$friends <- as.vector(G %*% rep(1, nrow(ca_pm25)))
ca_pm25$pm25_int <- as.vector(G %*% ca_pm25$pm25)
ca_pm25$pov_int <- as.vector(G %*% ca_pm25$pov)
ca_pm25$density_int <- as.vector(G %*% log(ca_pm25$density))

sortperm = order(ca_pm25$GEOID)
ca_pm25 = ca_pm25[sortperm, ]
G = G[sortperm, sortperm]

# What's the density of the graph?
sum(G != 0) / (nrow(G) * ncol(G))

set.seed(1234)

beta0 <- 0.05
beta1 <- 4
beta2 <- -0.4
beta0_int <- 0.1
beta1_int <- 8
beta2_int <- -0.8

# Generate spatial interference model
mu <-  beta0*ca_pm25$pm25 + 
  beta0_int*ca_pm25$pm25_int +  # interference term
  beta1*ca_pm25$pov +
  beta1_int*ca_pm25$pov_int +  # confounder of interference term
  beta2*log(ca_pm25$density) +
  beta2_int*ca_pm25$density_int + 5  # confounder of interference term
  
# Generate two types of outcomes: continuous and count
ca_pm25$Y1 <- mu + rnorm(nrow(ca_pm25),0,1)

```

Finally, we estimate the "modified treatment effect" (MTE) of increasing each unit's exposure by one. 
Confidence intervals are constructed using the classical asymptotic normal approximation, with
heteroscedasticity adjustment provided by the `sandwich` package. 
```{r}
Fbar = sum(G) / nrow(ca_pm25)
true_sample_mte = beta0 + beta0_int * Fbar

# Run linear glm, no interference
fit_lm <- lm(Y1 ~ pm25 + pov + log(density), data = ca_pm25)
mte = coef(fit_lm)["pm25"]
mte_ci = confint(fit_lm, vcov = vcovHC(fit_lm, "HC5"))["pm25", ]
mte_ci = c(mte, mte_ci)
mte_ci

# Run linear glm
fit_lm_int <- lm(Y1 ~ pm25 + pm25_int + pov + pov_int + log(density) + density_int + friends, data = ca_pm25)

# Compute estimate accounting for interference
mte_int = coef(fit_lm_int)["pm25"] + Fbar * coef(fit_lm_int)["pm25_int"]

# Compute its variance and confidence interval
V = vcovHC(fit_lm_int, "HC")
mte_var = V["pm25", "pm25"] + 
    (Fbar^2) * V["pm25_int", "pm25_int"] + 
    2 * Fbar * V["pm25", "pm25_int"]

mte_int_ci = mte_int + c("2.5%" = -1, "97.5%" = 1) * 1.96 * sqrt(mte_var)
mte_int_ci = c(mte_int, mte_int_ci)
mte_int_ci

```

The code below plots our results and produces an image file. 

```{r}
df_plot <- data.frame(
  method = factor(c("No Interference", "Interference"),
                  levels = c("Interference", "No Interference")),
  mean  = c(mte_ci[1], mte_int_ci[1]),
  lower = c(mte_ci[2], mte_int_ci[2]),
  upper = c(mte_ci[3], mte_int_ci[3])
)

ggplot(df_plot, aes(x = mean, y = method, color = method)) +
  geom_point(size = 3) +
  geom_errorbar(aes(xmin = lower, xmax = upper), width = 0.4, linewidth = 1) +
  geom_vline(xintercept = true_sample_mte, color = "darkgreen") +
  geom_text(aes(x=true_sample_mte+0.0025, label="True MTE", y=2), colour="darkgreen", angle=90, text=element_text(size=11)) +
  labs(
    x = expression("MTE of " * PM[2.5] * " on continuous outcome"),
    y = "",
    title = ""
  ) +
  scale_color_manual(values = c("Interference" = "#002e5c", "No Interference" = "#C67700")) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.title.x = element_text(margin = margin(t = 10))
  )

# Save as vector PDF for publication and PNG for web/preview
ggsave(filename = here("figs", "mte_pm25.pdf"),
  plot = last_plot(),
  device = "pdf",
  width = 5, height = 3, units = "in")

ggsave(filename = here("figs", "mte_pm25.png"),
  plot = last_plot(),
  width = 5, height = 3, units = "in", dpi = 300)
```