---
title: "Spatial confounding example"
output: html_document
---

```{r}
# Load necessary packages
library(sf)
library(mgcv)
library(gridExtra)
library(lme4)
library(stringr)
library(spdep)
library(INLA)
library(dplyr)
library(ggplot2)
library(viridis)
library(scales)
```

## Before starting

1) Ensure that you have a folder in this directory called "data"; if not, create it. 
2) Download the dataset from CAFE Dataverse (https://doi.org/10.7910/DVN/TN9JTT) as a csv and place it in the "data" folder. The file should be named "ca_bg_pm25_census_2016.csv".
3) Ensure that you have a folder in the parent directory spatial-cafe-webinar called "figs"; if not, create it. This is where the output figures will be saved.

## Import census block group shapefile, California 2016, for plotting
```{r importshapefile}
url <- "https://www2.census.gov/geo/tiger/TIGER2016/BG/tl_2016_06_bg.zip"
temp <- tempfile()
download.file(url, temp)
unzip(temp, exdir = "data/ca_bg_shapefile")
ca_bg_shp <- st_read("data/ca_bg_shapefile/tl_2016_06_bg.shp")
unlink(temp)
```

## Read in dataset
```{r readdata}
ca_pm25 <- read.csv('data/ca_bg_pm25_census_2016.csv')
# Convert GEOID to character using padding
ca_pm25$GEOID <- str_pad(ca_pm25$GEOID, width = 12, pad = "0")
```

## Generate synthetic outcome
```{r syntheticoutcome}
set.seed(1234)

beta0 <- 0.05
beta1 <- 0.25
beta2 <- 0.1
beta3 <- 0.5
beta4 <- -0.05

mu <-  beta0*ca_pm25$pm25 + 
  beta1*ca_pm25$pov +
  beta2*log(ca_pm25$density) +
  beta3*ca_pm25$ndvi +
  beta4*ca_pm25$D_min_km +
  rnorm(nrow(ca_pm25),0,0.5)
# Generate disease counts using population (ca_pm25$total_pop) and mu
ca_pm25$Y <- rpois(nrow(ca_pm25), exp(mu)/100 * ca_pm25$total_pop)
summary(ca_pm25$Y)
```

## Fit an oracle method (which adjusts for the unmeasured spatial confounders)
```{r oracle}
# Run glm poisson
fit_glm <- glm(Y ~ pm25 + pov + log(density) + ndvi + D_min_km,
               offset = log(total_pop),
               family = "poisson",
               data = ca_pm25)
summary(fit_glm) # coefficients look good
results <- data.frame(
  method = "oracle",
  estimate = summary(fit_glm)$coefficients["pm25", "Estimate"],
  conf_low = summary(fit_glm)$coefficients["pm25", "Estimate"] - 1.96 * summary(fit_glm)$coefficients["pm25", "Std. Error"],
  conf_high = summary(fit_glm)$coefficients["pm25", "Estimate"] + 1.96 * summary(fit_glm)$coefficients["pm25", "Std. Error"]
)
```

## Fit spatial method 
```{r spatial}
fit_spatial <- gam(
  Y ~ pm25 + # exposure
    pov + log(density) + # covariates
    s(cent_x, # x coordinate of block group centroid
      cent_y, # y coordinate of block group centroid
      k = 75, # basis dimension
      fx = T), # apply smoothing
  family = poisson(link = "log"), # Outcome is disease rate
  offset = log(total_pop), # offset of population
  data = ca_pm25,
  method = "GCV.Cp"
)
summary(fit_spatial)$p.coeff["pm25"]
results <- rbind.data.frame(
  results,
  data.frame(
    method = "spatial",
    estimate = summary(fit_spatial)$p.coeff["pm25"],
    conf_low = summary(fit_spatial)$p.coeff["pm25"] - 1.96 * summary(fit_spatial)$se["pm25"],
    conf_high = summary(fit_spatial)$p.coeff["pm25"] + 1.96 * summary(fit_spatial)$se["pm25"]
  )
)
```

## Spatial plus Step 1
```{r spatialplus1}
f_X_hat <- gam(pm25 ~ s(cent_x, # x coordinate of block group centroid
                        cent_y, # y coordinate of block group centroid
                        k = 75, # basis dimension
                        fx = T), # apply smoothing
               data = ca_pm25,
               method = 'GCV.Cp') # Generalized Cross Validation
ca_pm25$smoothed_pm25 <- f_X_hat$fitted.values
ca_pm25$residuals_pm25 <- f_X_hat$residuals
summary(f_X_hat)
```

## Spatial plus Step 2
```{r spatialplus2}
# Depending on the degrees of freedom k, this can take a long time to run
fit_spatial_plus <- mgcv::gam(
  Y ~ residuals_pm25 + # residuals of PM 2.5 from the previous regression
    pov + log(density) + # covariates
    s(cent_x, cent_y, k = 75, fx = T), # another thin plate spline of coordinates
  family = poisson(link = "log"), # Outcome is disease rate
  offset = log(total_pop), # offset of population
  data = ca_pm25,
  method = "GCV.Cp"
)
```

## Report results
```{r spatialplusreport}
print(paste("Estimate:", summary(fit_spatial_plus)$p.coeff["residuals_pm25"]))
```

## Save results
```{r spatialplussave}
summary(fit_spatial_plus)
# Save coefficient of residuals_pm25 and confidence interval in a data frame
results <- rbind.data.frame(
  results,
  data.frame(
    method = "spatial_plus",
    estimate = summary(fit_spatial_plus)$p.coeff["residuals_pm25"],
    conf_low = summary(fit_spatial_plus)$p.coeff["residuals_pm25"] - 1.96 * summary(fit_spatial_plus)$se["residuals_pm25"],
    conf_high = summary(fit_spatial_plus)$p.coeff["residuals_pm25"] + 1.96 * summary(fit_spatial_plus)$se["residuals_pm25"]
  )
)
```

## Plot PM2.5, the fitted values, and residuals
```{r plotspatialplus}
ca_bg_shp <- ca_bg_shp %>%
  left_join(st_drop_geometry(ca_pm25) %>% dplyr::select(GEOID, pm25, smoothed_pm25, residuals_pm25), by = "GEOID")
g1 <- ggplot() +
  geom_sf(data = ca_bg_shp, aes(fill = pm25), color = NA, linewidth = NA) +
  scale_fill_viridis_c(name = "PM2.5 (µg/m³)", na.value = "lightgrey") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_line(color = "transparent"),
    legend.background = element_rect(fill = alpha("white", 0.6)),
    legend.position = "bottom"
  )
g2 <- ggplot() +
  geom_sf(data = ca_bg_shp, aes(fill = smoothed_pm25), color = NA, linewidth = NA) +
  scale_fill_viridis_c(name = "Smoothed PM2.5 (µg/m³)", na.value = "lightgrey") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_line(color = "transparent"),
    legend.background = element_rect(fill = alpha("white", 0.6)),
    legend.position = "bottom"
  )
g3 <- ggplot() +
  geom_sf(data = ca_bg_shp, aes(fill = residuals_pm25), color = NA, linewidth = NA) +
  scale_fill_viridis_c(name = "Residuals (µg/m³)", na.value = "lightgrey") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_line(color = "transparent"),
    legend.background = element_rect(fill = alpha("white", 0.6)),
    legend.position = "bottom"
  )
png("../figs/ca_bg_pm25_spatialplus.png", width = 12, height = 5, units = "in", res = 300)
grid.arrange(g1, g2, g3, ncol = 3)
dev.off()
```

## Unadjusted GLM
```{r unadjustedglm}
# Compare to unadjusted glm
fit_ols <- glm(Y ~ pm25 + pov + log(density),
               offset = log(total_pop),
               family = "poisson",
               data = ca_pm25)
summary(fit_ols)
```

## Save unadjusted GLM results
```{r unadjustedglmsave}
results <- rbind.data.frame(
  results,
  data.frame(
    method = "unadjusted",
    estimate = summary(fit_ols)$coefficients["pm25", "Estimate"],
    conf_low = summary(fit_ols)$coefficients["pm25", "Estimate"] - 1.96 * summary(fit_ols)$coefficients["pm25", "Std. Error"],
    conf_high = summary(fit_ols)$coefficients["pm25", "Estimate"] + 1.96 * summary(fit_ols)$coefficients["pm25", "Std. Error"]
  )
)
```

## Plot the results
```{r plotresults}
# vertical line for true coef (beta0)
# Order results by Oracle, then Orindary GLM, then spatial, then spatial plus
results$method <- factor(results$method, levels = c("oracle", "unadjusted", "spatial", "spatial_plus"))
png("../figs/pm25_coefficients_comparison.png", width = 6, height = 4, units = "in", res = 350)
ggplot(results, aes(x = estimate, y = method)) +
  geom_point() +
  geom_errorbarh(aes(xmin = conf_low, xmax = conf_high), height = 0.2) +
  geom_vline(xintercept = beta0, linetype = "dashed", color = "red") +
  # add title
  #ggtitle("Estimated Coefficient for PM2.5") + # center title
  theme(plot.title = element_text(hjust = 0.5),
        title = "Estimated Coefficient for PM2.5") +
  xlab("") +
  ylab("") +
  theme_minimal()
dev.off()

```

## Fit a model with fixed effects at the county level
```{r countyfe}
# Extract county FIPS code from GEOID (first 5 digits)
ca_pm25$county_fips <- substr(ca_pm25$GEOID, 1, 5)
# Convert to factor
ca_pm25$county_fips <- as.factor(ca_pm25$county_fips)
fit_fe <- glm(Y ~ pm25 + pov + log(density) + county_fips,
              offset = log(total_pop),
              family = "poisson",
              data = ca_pm25)
results <- rbind.data.frame(
  results,
  data.frame(
    method = "county_fixed_effects",
    estimate = summary(fit_fe)$coefficients["pm25", "Estimate"],
    conf_low = summary(fit_fe)$coefficients["pm25", "Estimate"] - 1.96 * summary(fit_fe)$coefficients["pm25", "Std. Error"],
    conf_high = summary(fit_fe)$coefficients["pm25", "Estimate"] + 1.96 * summary(fit_fe)$coefficients["pm25", "Std. Error"]
  )
)
```

## Fit a model with random effects at the county level
```{r countyre}
fit_re <- glmer(Y ~ pm25 + pov + log(density) + (1|county_fips),
                offset = log(total_pop),
                family = "poisson",
                data = ca_pm25)
results <- rbind.data.frame(
  results,
  data.frame(
    method = "county_random_effects",
    estimate = summary(fit_re)$coefficients["pm25", "Estimate"],
    conf_low = summary(fit_re)$coefficients["pm25", "Estimate"] - 1.96 * summary(fit_re)$coefficients["pm25", "Std. Error"],
    conf_high = summary(fit_re)$coefficients["pm25", "Estimate"] + 1.96 * summary(fit_re)$coefficients["pm25", "Std. Error"]
  )
)
```

## Fit a model with random effects at the tract level
```{r tractre}
# Extract tract FIPS code from GEOID (first 11 digits)
ca_pm25$tract_fips <- substr(ca_pm25$GEOID, 1, 11)
# Convert to factor
ca_pm25$tract_fips <- as.factor(ca_pm25$tract_fips)

fit_re_tract <- glmer(Y ~ pm25 + pov + log(density) + (1|tract_fips),
                offset = log(total_pop),
                family = "poisson",
                data = ca_pm25)
results <- rbind.data.frame(
  results,
  data.frame(
    method = "tract_random_effects",
    estimate = summary(fit_re_tract)$coefficients["pm25", "Estimate"],
    conf_low = summary(fit_re_tract)$coefficients["pm25", "Estimate"] - 1.96 * summary(fit_re_tract)$coefficients["pm25", "Std. Error"],
    conf_high = summary(fit_re_tract)$coefficients["pm25", "Estimate"] + 1.96 * summary(fit_re_tract)$coefficients["pm25", "Std. Error"]
  )
)
```

## Fit a conditional autoregressive model
```{r car}
# Subset ca_pm25 to observations with nonmissing and finite Y, pm25, pov, log(density)
ca_pm25_complete <- ca_pm25 %>%
  filter(!is.na(Y) & is.finite(Y) &
           !is.na(pm25) & is.finite(pm25) &
           !is.na(pov) & is.finite(pov) &
           !is.na(density) & is.finite(density))
# Change GEOID to character using padding
ca_pm25_complete$GEOID <- str_pad(as.character(ca_pm25_complete$GEOID), width = 12, pad = "0")
# Subset ca_bg_shp to GEOIDs in ca_pm25_complete
ca_bg_shp <- st_read("data/ca_bg_shapefile/tl_2016_06_bg.shp") # reset 
ca_bg_shp <- ca_bg_shp %>%
  filter(GEOID %in% ca_pm25_complete$GEOID)
bg <- st_make_valid(ca_bg_shp)

nb <- poly2nb(bg, queen = TRUE)
W  <- nb2mat(nb, style = "B", zero.policy = TRUE)  # 0/1 adjacency matrix
spdep::nb2INLA(file = "bg.graph", nb = nb)
g <- INLA::inla.read.graph("bg.graph")

bg$id <- seq_len(nrow(bg))  

# Assign bg pm25, pov, density, total_pop, Y to st_drop_geometry(bg) merging on GEOID
bg <- bg %>%
  left_join(
    ca_pm25_complete %>%
      dplyr::select(GEOID, pm25, pov, density, total_pop, Y),
    by = "GEOID"
  )
fit_bym2 <- inla(
  Y ~ pm25 + pov + log(density) +
    f(id, model = "bym2", graph = g, scale.model = TRUE),
  family = "poisson",   
  data = st_drop_geometry(bg),
  control.predictor = list(compute = TRUE)
)
summary(fit_bym2)
results <- rbind.data.frame(
  results,
  data.frame(
    method = "BYM2",
    estimate = fit_bym2$summary.fixed["pm25", "mean"],
    conf_low = fit_bym2$summary.fixed["pm25", "0.025quant"],
    conf_high = fit_bym2$summary.fixed["pm25", "0.975quant"]
  )
)
```

## Plot all results on a single plot
```{r plotallresults}
# Order results by Oracle, then Orindary GLM, then spatial, then spatial plus, then county fe, county re, tract re, CAR
lvl <- c("oracle", "unadjusted", "spatial", "spatial_plus",
         "county_fixed_effects", "county_random_effects",
         "tract_random_effects", "BYM2")

results <- results %>%
  mutate(method = factor(method, levels = lvl))

png("../figs/pm25_coefficients_comparison_all.png", width = 6, height = 4, units = "in", res = 350)
ggplot(results, aes(x = estimate, y = method)) +
  geom_point() +
  geom_errorbarh(aes(xmin = conf_low, xmax = conf_high), height = 0.2) +
  geom_vline(xintercept = beta0, linetype = "dashed", color = "red") +
  scale_y_discrete(limits = rev(lvl)) +   # <- puts oracle on top, BYM2 on bottom
  theme_minimal() +
  xlab("") + ylab("")
dev.off()

```