---
title: "spatial_confounding_interference"
output: html_document
---

```{r}
# Load necessary packages
library(tidyverse)
library(here)
library(igraph)
library(tidycensus)
library(sf)
library(lmtest)
library(sandwich)
library(stargazer)
library(terra)
library(exactextractr)
library(mgcv)
library(gridExtra)
library(lme4)
library(stringr)
library(spdep)
library(INLA)

options(tigris_use_cache = T)
```

## Import census block group shapefile, California 2016

```{r}
url <- "https://www2.census.gov/geo/tiger/TIGER2016/BG/tl_2016_06_bg.zip"
temp <- tempfile()
download.file(url, temp)
unzip(temp, exdir = "data/ca_bg_shapefile")
ca_bg_shp <- st_read("data/ca_bg_shapefile/tl_2016_06_bg.shp")
unlink(temp)
```

## Read in 1km x 1km PM2.5 data from CAFE Dataverse, Di et al 2019
See https://doi.org/10.7910/DVN/58C6HG for more information.
```{r}
# Query from the Dataverse API
url <- "https://dataverse.harvard.edu/api/access/datafile/11072770"
download.file(url, destfile = "data/annual_pm25.zip")
unzip("data/annual_pm25.zip", exdir = "data/annual_pm25")
pm25_1km_2016 <- readRDS("data/annual_pm25/annual-rds/PM25-2016.rds")

grid_url <- "https://dataverse.harvard.edu/api/access/datafile/11071648"
download.file(grid_url, destfile = "data/grid.csv")
grid <- read.csv("data/grid.csv")
# Assign PM to grid
grid$pm <- pm25_1km_2016
```

## Aggregate PM2.5 to block group level
Naively, let's just take the average of PM values of all grid cells that intersect each block group.
```{r}
ca_bg_ll <- st_transform(ca_bg_shp, 4326)
bb <- st_bbox(ca_bg_ll)

grid_ca <- grid[
  grid$lon >= bb["xmin"] & grid$lon <= bb["xmax"] &
  grid$lat >= bb["ymin"] & grid$lat <= bb["ymax"],
]

grid_sf_ll <- st_as_sf(grid_ca, coords = c("lon", "lat"), crs = 4326)

crs_m <- 3310  # CA Albers coordinate system (meters)
grid_sf <- st_transform(grid_sf_ll, crs_m)
ca_bg  <- st_transform(ca_bg_shp, crs_m)

# rasterize
pts <- terra::vect(grid_sf)

r <- terra::rast(terra::ext(pts), res = 1000, crs = terra::crs(pts))  # 1km cells
pm_r <- terra::rasterize(pts, r, field = "pm", fun = mean, na.rm = TRUE)

# block group polygon means
ca_bg_pm25 <- ca_bg[, "GEOID"]
ca_bg_pm25$pm25 <- exactextractr::exact_extract(pm_r, ca_bg_pm25, "mean")

summary(ca_bg_pm25$pm25)

# Plot as image 
png("../figs/ca_bg_pm25.png", width = 6, height = 5, units = "in", res = 300)
ggplot() +
  geom_sf(data = ca_bg_pm25, aes(fill = pm25), color = NA) +
  scale_fill_viridis_c(name = "PM2.5 (µg/m³)", na.value = "lightgrey") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_line(color = "transparent"),
    legend.background = element_rect(fill = alpha("white", 0.6))
  )
dev.off()

```

## Extract ACS 5 Year Census Data from 2012-2016
```{r}
# Set Census API key before proceeding
# Variable names
vars <- c(
  # income / poverty
  med_hh_income = "B19013_001", # in last year's inflation-adjusted dollars
  med_home_value = "B25077_001", # 
  pov_den = "B17010_001", pov_num = "B17010_002", # POVERTY STATUS IN THE PAST 12 MONTHS
  # employment
  emp_den     = "B23025_001", emp_num    = "B23025_002", # EMPLOYMENT STATUS FOR THE POPULATION 16 YEARS AND OVER
  # education (25+)
  HS_diploma    = "B15003_017", HS_equivalent = "B15003_018",
  some_college  = "B15003_019", some_college_plus = "B15003_020",
  assoc        = "B15003_021",
  bach          = "B15003_022", master       = "B15003_023",
  prof          = "B15003_024", doct         = "B15003_025", edu_den = "B15003_001",
  # tenure
  owner_occ     = "B25003_002", tenure_den   = "B25003_001", # TENURE 
  # race / ethnicity
  black         = "B02001_003", race_den     = "B02001_001",   # Black alone / total
  hisp          = "B03003_003", hisp_den     = "B03003_001",   # Hispanic / total
  # total population
  total_pop     = "B01003_001", 
  # median year structure built
  med_year_built = "B25035_001",
  # age < 5
  age_denom = "B01001_001", age_u5_male = "B01001_003", age_u5_female = "B01001_027",
  # age > 65
  age_6566_male = "B01001_020", age_6769_male = "B01001_021", age_7074_male = "B01001_022",
  age_7579_male = "B01001_023", age_8084_male = "B01001_024", age_85up_male = "B01001_025",
  age_6566_female = "B01001_044", age_6769_female = "B01001_045", age_7074_female = "B01001_046",
  age_7579_female = "B01001_047", age_8084_female = "B01001_048", age_85up_female = "B01001_049"
)

expected_E_cols <- paste0(names(vars), "E")

pull_one_year <- function(y) {
  # Only request variables that are available
  lv <- load_variables(y, "acs5", cache = TRUE)
  present <- names(vars)[vars %in% lv$name]
  vars_y  <- vars[present]
  if (length(vars_y) == 0) return(tibble())
  
  df <- get_acs(
    geography = "block group",
    state = "CA", 
    variables = vars_y,
    year = y, survey = "acs5",
    cache_table = TRUE, geometry = TRUE, output = "wide"
  )
  
  # Back-fill any missing expected columns with NA 
  for (col in setdiff(expected_E_cols, names(df))) df[[col]] <- NA_real_
  
  df %>%
    transmute(
      GEOID, NAME, year = y,
      
      med_hh_income = med_hh_incomeE,
      med_home_value = med_home_valueE,
      pov_rate      = ifelse(pov_denE     > 0, pov_numE / pov_denE, NA_real_),
      emp_rate    = ifelse(emp_denE   > 0, emp_numE   / emp_denE,   NA_real_),
      
      grad_HS     = ifelse(edu_denE     > 0, (HS_diplomaE + HS_equivalentE + some_collegeE + some_college_plusE + assocE +
                                                bachE + masterE + profE + doctE) / edu_denE, NA_real_),
      
      homeown_rate  = ifelse(tenure_denE  > 0, owner_occE   / tenure_denE,  NA_real_),
      
      pct_black     = ifelse(race_denE    > 0, blackE       / race_denE,    NA_real_),
      pct_hispanic  = ifelse(hisp_denE    > 0, hispE        / hisp_denE,    NA_real_),
      total_pop = total_popE,
      med_year_built = med_year_builtE,
      pct_under5 = ifelse(age_denomE > 0, (age_u5_maleE + age_u5_femaleE) / age_denomE, NA_real_),
      pct_over65 = ifelse(age_denomE > 0, (age_6566_maleE + age_6769_maleE + age_7074_maleE +
                                             age_7579_maleE + age_8084_maleE + age_85up_maleE +
                                             age_6566_femaleE + age_6769_femaleE + age_7074_femaleE +
                                              age_7579_femaleE + age_8084_femaleE + age_85up_femaleE) / age_denomE, NA_real_)
      
    )
}
# Pull just 2016 california
ca_census_2016 <- pull_one_year(2016)
```

## Merge the census data with the PM2.5 data
```{r}
mean(ca_census_2016$GEOID %in% ca_bg_pm25$GEOID) # 1
mean(ca_bg_pm25$GEOID %in% ca_census_2016$GEOID) # 1
ca_pm25 <- merge(ca_bg_pm25, st_drop_geometry(ca_census_2016), by = "GEOID")
ca_pm25 <- ca_pm25 %>%
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    pov = pov_rate,
    density = total_pop / area_km2
  )
# Get centroids of each block group and save coordinates as well
ca_pm25 <- ca_pm25 %>%
  mutate(
    centroid = st_centroid(geometry),
    cent_x = st_coordinates(centroid)[,1],
    cent_y = st_coordinates(centroid)[,2]
  ) %>%
  select(-centroid)

summary(ca_pm25)
```



## Read in health center data
```{r}
url <- "https://data.chhs.ca.gov/dataset/3b5b80e8-6b8d-4715-b3c0-2699af6e72e5/resource/f0ae5731-fef8-417f-839d-54a0ed3a126e/download/health_facility_locations.csv"
download.file(url, destfile = "data/health_facility_locations.csv")
health <- read.csv('data/health_facility_locations.csv')
health_sf <- st_as_sf(health, coords = c("LONGITUDE", "LATITUDE"), crs = 4326)
health_sf <- st_transform(health_sf, 3310) # CA Albers (meters)

ca_pm25_sf <- st_as_sf(ca_pm25, coords = c("cent_x", "cent_y"), 
                       crs = 3310)
# Get distances to nearest health facility for each block group centroid
# D <- st_distance(ca_pm25_sf, health_sf)
# ca_pm25$D_min_m <- apply(D, 1, min)
idx <- st_nearest_feature(ca_pm25_sf, health_sf)
ca_pm25$D_min_m <- st_distance(
  ca_pm25_sf,
  health_sf[idx, ],
  by_element = TRUE
)
# Convert to km
ca_pm25$D_min_km <- ca_pm25$D_min_m / 1000
summary(ca_pm25$D_min_km)
# Convert units to numeric
ca_pm25$D_min_km <- as.numeric(ca_pm25$D_min_km)
# Convert ca_pm25$GEOID to character, padding with 0 at front
ca_pm25$GEOID <- str_pad(as.character(ca_pm25$GEOID), width = 12, pad = "0")
# Merge with ca_pm25
ca_bg_shp <- ca_bg_shp %>%
  left_join(st_drop_geometry(ca_pm25) %>%
              dplyr::select(D_min_km, GEOID), by = "GEOID")

# Plot distance to nearest health facility
png('../figs/distance_to_health_facility_ca.png', width = 6, 
    height = 5, units = 'in', res = 300)
ggplot() + 
  geom_sf(data = ca_bg_shp, aes(fill = D_min_km), color = NA) +
  scale_fill_viridis_c(name = "km to nearest\nCA health facility", na.value = "lightgrey", option = 'magma') +
  theme_minimal() + # remove gridlines and background
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_line(color = "transparent"),
    legend.background = element_rect(fill = alpha("white", 0.6)),
    legend.position = "right"
  )
dev.off()
```

## Read in NDVI data
```{r}
# https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/V0BDE7
ndvi <- read.csv('data/ndvi_seasonal_zip20002016.csv')
ndvi <- ndvi %>%
  filter(year == 2016) %>%
  select(ZIP, NDVI_summer)
# Convert ndvi zip to five digit character padding with zeros
ndvi$ZIP <- str_pad(as.character(ndvi$ZIP), width = 5, pad = "0")

# zip to zcta crosswalk
# https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/HYNJSZ
xwalk <- read.csv('data/zip2zcta_master_xwalk.csv')
# Convert both zip and zcta to 5 digit characters with padding
xwalk$zip <- str_pad(as.character(xwalk$zip), width = 5, pad = "0")
xwalk$zcta <- str_pad(as.character(xwalk$zcta), width = 5, pad = "0")
# Create zcta as column in ndvi by merging with xwalk
ndvi <- ndvi %>%
  left_join(xwalk %>% dplyr::select(zip, zcta), by = c("ZIP" = "zip")) %>%
  select(ZIP, zcta, NDVI_summer)

# Read in zip code shapefile
url <- "https://www2.census.gov/geo/tiger/TIGER2016/ZCTA5/tl_2016_us_zcta510.zip"
temp <- tempfile()
download.file(url, temp)
unzip(temp, exdir = "data/")
zip_shp <- st_read("data/tl_2016_us_zcta510.shp")
# Merge ndvi with zip_shp by matching ndvi$zcta with zip_shp$ZCTA5CE10
zip_shp <- zip_shp %>%
  left_join(ndvi, by = c("ZCTA5CE10" = "zcta"))

# Plot ndvi in California 
st_crs(zip_shp) == st_crs(ca_bg_shp) # TRUE, both are in NAD83
# Assign NDVI_summer to ca_bg_shp by spatial aggregation
bb <- st_as_sfc(st_bbox(ca_bg_shp))
zip_ca <- zip_shp[ lengths(st_intersects(zip_shp, bb)) > 0, c("NDVI_summer") ]

# 2) Build intersection index (list of zip indices per bg)
idx <- st_intersects(ca_bg_shp, zip_ca)  # sparse list, fast

# 3) Compute mean NDVI for each bg (or GEOID) without expanding a join
bg_ndvi <- vapply(idx, function(ii) mean(zip_ca$NDVI_summer[ii], na.rm = TRUE), numeric(1))

ca_bg_shp$NDVI_summer <- bg_ndvi
summary(bg_ndvi)

png('../figs/ndvi_summer_2016.png', width = 6, height = 5, units = 'in', res = 300)
ggplot() +
  geom_sf(data = ca_bg_shp, aes(fill = NDVI_summer), color = NA) + # plot from white to green 0 to 1
  scale_fill_gradient(name = "NDVI (Summer 2016)", low = "white", high = "darkgreen", na.value = "lightgrey") +
  #scale_fill_viridis_c(name = "NDVI (Summer 2016)", na.value = "lightgrey", option = 'mako') +
  #scale_fill_viridis_c(name = "NDVI (Summer 2016)", na.value = "lightgrey") +
  theme_minimal() + # remove gridlines and
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_line(color = "transparent"),
    legend.background = element_rect(fill = alpha("white", 0.6)),
    legend.position = "right"
  )
dev.off()

# Merge with ca_pm25
ca_pm25 <- ca_pm25 %>%
  left_join(st_drop_geometry(ca_bg_shp) %>% dplyr::select(GEOID, NDVI_summer), by = "GEOID") %>%
  rename(ndvi = NDVI_summer)
```

## Print five rows at random
```{r}
summary(ca_pm25)
write.csv(st_drop_geometry(ca_pm25), "data/ca_bg_pm25_census_2016.csv", row.names = F)

# Sample 5 rows at random, distance, ndvi, PM2.5, poverty, density
# Round distance, ndvi, pm, poverty density to 2 decimal places
set.seed(122)
print(ca_pm25 %>%
  st_drop_geometry %>%
  select(GEOID, D_min_km, ndvi, pov, density, pm25) %>% 
  mutate(
    D_min_km = round(D_min_km, 2),
    ndvi = round(ndvi, 2),
    pov = round(pov, 2),
    density = round(density, 2),
    pm25 = round(pm25, 2)
  ) %>%
  sample_n(5), row.names = F)
```



## Generate synthetic outcome
```{r}
set.seed(1234)

beta0 <- 0.05
beta1 <- 0.25
beta2 <- 0.1
beta3 <- 0.3
beta4 <- -0.05

mu <-  beta0*ca_pm25$pm25 + 
  beta1*ca_pm25$pov +
  beta2*log(ca_pm25$density) +
  beta3*ca_pm25$ndvi +
  beta4*ca_pm25$D_min_km +
  rnorm(nrow(ca_pm25),0,0.5)
# Generate disease counts using population (ca_pm25$total_pop) and mu
ca_pm25$Y <- rpois(nrow(ca_pm25), exp(mu)/500 * ca_pm25$total_pop)
summary(ca_pm25$Y)

# Run glm poisson
fit_glm <- glm(Y ~ pm25 + pov + log(density) + ndvi + D_min_km,
               offset = log(total_pop),
               family = "poisson",
               data = ca_pm25)
summary(fit_glm) # ORACLE: coefficients look good
results <- data.frame(
  method = "oracle",
  estimate = summary(fit_glm)$coefficients["pm25", "Estimate"],
  conf_low = summary(fit_glm)$coefficients["pm25", "Estimate"] - 1.96 * summary(fit_glm)$coefficients["pm25", "Std. Error"],
  conf_high = summary(fit_glm)$coefficients["pm25", "Estimate"] + 1.96 * summary(fit_glm)$coefficients["pm25", "Std. Error"]
)
```

## Spatial method 
```{r}
fit_spatial <- mgcv::gam(
  Y ~ pm25 + # exposure
    pov + log(density) + # covariates
    s(cent_x, # x coordinate of block group centroid
      cent_y, # y coordinate of block group centroid
      k = 100, # basis dimension
      fx = T), # apply smoothing
  family = poisson(link = "log"), # Outcome is disease rate
  offset = log(total_pop), # offset of population
  data = ca_pm25,
  method = "GCV.Cp"
)
```

## Save spatial results
```{r}
results <- rbind.data.frame(
  results,
  data.frame(
    method = "spatial",
    estimate = summary(fit_spatial)$p.coeff["pm25"],
    conf_low = summary(fit_spatial)$p.coeff["pm25"] - 1.96 * summary(fit_spatial)$se["pm25"],
    conf_high = summary(fit_spatial)$p.coeff["pm25"] + 1.96 * summary(fit_spatial)$se["pm25"]
  )
)
```

## Spatial plus Step 1
```{r}
f_X_hat <- gam(pm25 ~ s(cent_x, # x coordinate of block group centroid
                        cent_y, # y coordinate of block group centroid
                        k = 100, # basis dimension
                        fx = T), # apply smoothing
               data = ca_pm25,
               method = 'GCV.Cp') # Generalized Cross Validation
ca_pm25$smoothed_pm25 <- f_X_hat$fitted.values
ca_pm25$residuals_pm25 <- f_X_hat$residuals
summary(f_X_hat)
```

## Spatial plus Step 2
```{r}
fit_spatial_plus <- mgcv::gam(
  Y ~ residuals_pm25 + # residuals of PM 2.5 from the previous regression
    pov + log(density) + # covariates
    s(cent_x, cent_y, k = 100, fx = T), # another thin plate spline of coordinates
  family = poisson(link = "log"), # Outcome is disease rate
  offset = log(total_pop), # offset of population
  data = ca_pm25,
  method = "GCV.Cp"
)
```

## Report results
```{r}
print(paste("Estimate:", summary(fit_spatial_plus)$p.coeff["residuals_pm25"]))
```

## Save results
```{r}
summary(fit_spatial_plus)
# Save coefficient of residuals_pm25 and confidence interval in a data frame
results <- rbind.data.frame(
  results,
  data.frame(
    method = "spatial_plus",
    estimate = summary(fit_spatial_plus)$p.coeff["residuals_pm25"],
    conf_low = summary(fit_spatial_plus)$p.coeff["residuals_pm25"] - 1.96 * summary(fit_spatial_plus)$se["residuals_pm25"],
    conf_high = summary(fit_spatial_plus)$p.coeff["residuals_pm25"] + 1.96 * summary(fit_spatial_plus)$se["residuals_pm25"]
  )
)
```

## Plot PM2.5, the fitted values, and residuals
```{r}
ca_bg_shp <- ca_bg_shp %>%
  left_join(st_drop_geometry(ca_pm25) %>% dplyr::select(GEOID, pm25, smoothed_pm25, residuals_pm25), by = "GEOID")
g1 <- ggplot() +
  geom_sf(data = ca_bg_shp, aes(fill = pm25), color = NA, linewidth = NA) +
  scale_fill_viridis_c(name = "PM2.5 (µg/m³)", na.value = "lightgrey") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_line(color = "transparent"),
    legend.background = element_rect(fill = alpha("white", 0.6)),
    legend.position = "bottom"
  )
g2 <- ggplot() +
  geom_sf(data = ca_bg_shp, aes(fill = smoothed_pm25), color = NA, linewidth = NA) +
  scale_fill_viridis_c(name = "Smoothed PM2.5 (µg/m³)", na.value = "lightgrey") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_line(color = "transparent"),
    legend.background = element_rect(fill = alpha("white", 0.6)),
    legend.position = "bottom"
  )
g3 <- ggplot() +
  geom_sf(data = ca_bg_shp, aes(fill = residuals_pm25), color = NA, linewidth = NA) +
  scale_fill_viridis_c(name = "Residuals (µg/m³)", na.value = "lightgrey") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_line(color = "transparent"),
    legend.background = element_rect(fill = alpha("white", 0.6)),
    legend.position = "bottom"
  )
png("../figs/ca_bg_pm25_spatialplus.png", width = 12, height = 5, units = "in", res = 300)
grid.arrange(g1, g2, g3, ncol = 3)
dev.off()
```

## Ordinary GLM
```{r}
# Compare to ordinary glm
fit_ols <- glm(Y ~ pm25 + pov + log(density),
               offset = log(total_pop),
               family = "poisson",
               data = ca_pm25)
summary(fit_ols)
```

## Save Ordinary GLM results
```{r}
results <- rbind.data.frame(
  results,
  data.frame(
    method = "ordinary_glm",
    estimate = summary(fit_ols)$coefficients["pm25", "Estimate"],
    conf_low = summary(fit_ols)$coefficients["pm25", "Estimate"] - 1.96 * summary(fit_ols)$coefficients["pm25", "Std. Error"],
    conf_high = summary(fit_ols)$coefficients["pm25", "Estimate"] + 1.96 * summary(fit_ols)$coefficients["pm25", "Std. Error"]
  )
)
```

## Plot the results
```{r}
# vertical line for true coef (beta0)
# Order results by Oracle, then Orindary GLM, then spatial, then spatial plus
results$method <- factor(results$method, levels = c("oracle", "ordinary_glm", "spatial", "spatial_plus"))
# Rename ordinary_glm as unadjusted
results$method <- recode(results$method, ordinary_glm = "unadjusted")
png("../figs/pm25_coefficients_comparison.png", width = 6, height = 4, units = "in", res = 350)
ggplot(results, aes(x = estimate, y = method)) +
  geom_point() +
  geom_errorbarh(aes(xmin = conf_low, xmax = conf_high), height = 0.2) +
  geom_vline(xintercept = beta0, linetype = "dashed", color = "red") +
  # add title
  #ggtitle("Estimated Coefficient for PM2.5") + # center title
  theme(plot.title = element_text(hjust = 0.5),
        title = "Estimated Coefficient for PM2.5") +
  xlab("") +
  ylab("") +
  theme_minimal()
dev.off()

```

## Fit a model with fixed effects at the county level
```{r}
# Extract county FIPS code from GEOID (first 5 digits)
ca_pm25$county_fips <- substr(ca_pm25$GEOID, 1, 5)
# Convert to factor
ca_pm25$county_fips <- as.factor(ca_pm25$county_fips)
fit_fe <- glm(Y ~ pm25 + pov + log(density) + county_fips,
              offset = log(total_pop),
              family = "poisson",
              data = ca_pm25)
results <- rbind.data.frame(
  results,
  data.frame(
    method = "county_fixed_effects",
    estimate = summary(fit_fe)$coefficients["pm25", "Estimate"],
    conf_low = summary(fit_fe)$coefficients["pm25", "Estimate"] - 1.96 * summary(fit_fe)$coefficients["pm25", "Std. Error"],
    conf_high = summary(fit_fe)$coefficients["pm25", "Estimate"] + 1.96 * summary(fit_fe)$coefficients["pm25", "Std. Error"]
  )
)
```

## Fit a model with random effects at the county level
```{r}
fit_re <- glmer(Y ~ pm25 + pov + log(density) + (1|county_fips),
                offset = log(total_pop),
                family = "poisson",
                data = ca_pm25)
results <- rbind.data.frame(
  results,
  data.frame(
    method = "county_random_effects",
    estimate = summary(fit_re)$coefficients["pm25", "Estimate"],
    conf_low = summary(fit_re)$coefficients["pm25", "Estimate"] - 1.96 * summary(fit_re)$coefficients["pm25", "Std. Error"],
    conf_high = summary(fit_re)$coefficients["pm25", "Estimate"] + 1.96 * summary(fit_re)$coefficients["pm25", "Std. Error"]
  )
)
```

## Fit a model with fixed effects at tract level
```{r}
# Extract tract FIPS code from GEOID (first 11 digits)
ca_pm25$tract_fips <- substr(ca_pm25$GEOID, 1, 11)
# Convert to factor
ca_pm25$tract_fips <- as.factor(ca_pm25$tract_fips)
fit_fe_tract <- glm(Y ~ pm25 + pov + log(density) + tract_fips,
              offset = log(total_pop),
              family = "poisson",
              data = ca_pm25)
results <- rbind.data.frame(
  results,
  data.frame(
    method = "tract_fixed_effects",
    estimate = summary(fit_fe_tract)$coefficients["pm25", "Estimate"],
    conf_low = summary(fit_fe_tract)$coefficients["pm25", "Estimate"] - 1.96 * summary(fit_fe_tract)$coefficients["pm25", "Std. Error"],
    conf_high = summary(fit_fe_tract)$coefficients["pm25", "Estimate"] + 1.96 * summary(fit_fe_tract)$coefficients["pm25", "Std. Error"]
  )
)
```

## Fit a model with random effects at the tract level
```{r}
fit_re_tract <- glmer(Y ~ pm25 + pov + log(density) + (1|tract_fips),
                offset = log(total_pop),
                family = "poisson",
                data = ca_pm25)
results <- rbind.data.frame(
  results,
  data.frame(
    method = "tract_random_effects",
    estimate = summary(fit_re_tract)$coefficients["pm25", "Estimate"],
    conf_low = summary(fit_re_tract)$coefficients["pm25", "Estimate"] - 1.96 * summary(fit_re_tract)$coefficients["pm25", "Std. Error"],
    conf_high = summary(fit_re_tract)$coefficients["pm25", "Estimate"] + 1.96 * summary(fit_re_tract)$coefficients["pm25", "Std. Error"]
  )
)
```

## Fit a conditional autoregressive model
```{r}
# Subset ca_pm25 to observations with nonmissing and finite Y, pm25, pov, log(density)
ca_pm25_complete <- ca_pm25 %>%
  filter(!is.na(Y) & is.finite(Y) &
           !is.na(pm25) & is.finite(pm25) &
           !is.na(pov) & is.finite(pov) &
           !is.na(density) & is.finite(density))
# Change GEOID to character using padding
ca_pm25_complete$GEOID <- str_pad(as.character(ca_pm25_complete$GEOID), width = 12, pad = "0")
# Subset ca_bg_shp to GEOIDs in ca_pm25_complete
ca_bg_shp <- ca_bg_shp %>%
  filter(GEOID %in% ca_pm25_complete$GEOID)
bg <- st_make_valid(ca_bg_shp)

nb <- poly2nb(bg, queen = TRUE)
W  <- nb2mat(nb, style = "B", zero.policy = TRUE)  # 0/1 adjacency matrix
spdep::nb2INLA(file = "bg.graph", nb = nb)
g <- INLA::inla.read.graph("bg.graph")

bg$id <- seq_len(nrow(bg))  # integer id 1..N

# Assign bg pm25, pov, density, total_pop, Y to st_drop_geometry(bg) merging on GEOID
bg <- bg %>%
  left_join(
    ca_pm25_complete %>%
      dplyr::select(GEOID, pm25, pov, density, total_pop, Y),
    by = "GEOID"
  )
fit_bym2 <- inla(
  Y ~ pm25 + pov + log(density) +
    f(id, model = "bym2", graph = g, scale.model = TRUE),
  family = "poisson",   
  data = st_drop_geometry(bg),
  control.predictor = list(compute = TRUE)
)
summary(fit_bym2)
results <- rbind.data.frame(
  results,
  data.frame(
    method = "BYM2",
    estimate = fit_bym2$summary.fixed["pm25", "mean"],
    conf_low = fit_bym2$summary.fixed["pm25", "0.025quant"],
    conf_high = fit_bym2$summary.fixed["pm25", "0.975quant"]
  )
)
```

## Plot results similar to above
```{r}
# Order results by Oracle, then Orindary GLM, then spatial, then spatial plus, then county fe, county re, tract re, CAR
lvl <- c("oracle", "unadjusted", "spatial", "spatial_plus",
         "county_fixed_effects", "county_random_effects",
         "tract_random_effects", "BYM2")

results <- results %>%
  mutate(method = factor(method, levels = lvl))

png("../figs/pm25_coefficients_comparison_all.png", width = 6, height = 4, units = "in", res = 350)
ggplot(results, aes(x = estimate, y = method)) +
  geom_point() +
  geom_errorbarh(aes(xmin = conf_low, xmax = conf_high), height = 0.2) +
  geom_vline(xintercept = beta0, linetype = "dashed", color = "red") +
  scale_y_discrete(limits = rev(lvl)) +   # <- puts oracle on top, BYM2 on bottom
  theme_minimal() +
  xlab("") + ylab("")
dev.off()

```

## Diagnosing spatial confounding
```{r}
# let's extract the residuals from ordinary glm and see if they are spatially structured
ca_pm25_complete$residuals_ols <- resid(fit_ols)
coords <- cbind(ca_pm25_complete$cent_x,
                ca_pm25_complete$cent_y)
knn <- knearneigh(coords, k = 8)  # k = 5–10 is typical
nb  <- knn2nb(knn)
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)
moran.test(
  ca_pm25_complete$residuals_ols,
  lw,
  zero.policy = TRUE
)

# Calculate MOran's I for residuals_ols using the adjacency matrix W
moran_ols <- moran.test(ca_pm25_complete$residuals_ols, listw = mat2listw(W), zero.policy = TRUE)
print(moran_ols)
# Merge with ca_bg_shp
ca_bg_shp <- st_read("data/ca_bg_shapefile/tl_2016_06_bg.shp")
ca_bg_shp <- ca_bg_shp %>%
  left_join(st_drop_geometry(ca_pm25_complete) %>% dplyr::select(GEOID, residuals_ols), by = "GEOID")

png("../figs/residuals_ols.png", width = 6, height = 5, units = "in", res = 300)
 ggplot() +
  geom_sf(data = ca_bg_shp, aes(fill = residuals_ols), color = NA, linewidth = NA) +
  scale_fill_viridis_c(name = "Residuals from non-spatial GLM", na.value = "lightgrey", limits = c(-11,11)) +
  # scale_fill_gradient2(name = "Residuals from non-spatial GLM", low = "lightblue", mid = "lightyellow", high = "red", na.value = "lightgrey",
  #                      breaks = c(-10, -5, 0, 5, 10), limits = c(-6,6)
  #                      ) +
  theme_minimal() + #
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_line(color = "transparent"),
    legend.background = element_rect(fill = alpha("white", 0.6)),
    legend.position = "right"
  )
dev.off()
```








