---
title: "spatial_confounding_interference"
output: html_document
---

```{r}
# Load necessary packages
library(tidyverse)
library(here)
library(igraph)
library(tidycensus)
library(sf)
library(lmtest)
library(sandwich)
library(stargazer)
library(terra)
library(exactextractr)
library(mgcv)
library(gridExtra)
options(tigris_use_cache = T)
```

## Import census block group shapefile, California 2016

```{r}
url <- "https://www2.census.gov/geo/tiger/TIGER2016/BG/tl_2016_06_bg.zip"
temp <- tempfile()
download.file(url, temp)
unzip(temp, exdir = "data/ca_bg_shapefile")
ca_bg_shp <- st_read("data/ca_bg_shapefile/tl_2016_06_bg.shp")
unlink(temp)
```

## Read in 1km x 1km PM2.5 data from CAFE Dataverse, Di et al 2019
See https://doi.org/10.7910/DVN/58C6HG for more information.
```{r}
# Query from the Dataverse API
url <- "https://dataverse.harvard.edu/api/access/datafile/11072770"
download.file(url, destfile = "data/annual_pm25.zip")
unzip("data/annual_pm25.zip", exdir = "data/annual_pm25")
pm25_1km_2016 <- readRDS("data/annual_pm25/annual-rds/PM25-2016.rds")

grid_url <- "https://dataverse.harvard.edu/api/access/datafile/11071648"
download.file(grid_url, destfile = "data/grid.csv")
grid <- read.csv("data/grid.csv")
# Assign PM to grid
grid$pm <- pm25_1km_2016
```

## Aggregate PM2.5 to block group level
Naively, let's just take the average of PM values of all grid cells that intersect each block group.
```{r}
ca_bg_ll <- st_transform(ca_bg_shp, 4326)
bb <- st_bbox(ca_bg_ll)

grid_ca <- grid[
  grid$lon >= bb["xmin"] & grid$lon <= bb["xmax"] &
  grid$lat >= bb["ymin"] & grid$lat <= bb["ymax"],
]

grid_sf_ll <- st_as_sf(grid_ca, coords = c("lon", "lat"), crs = 4326)

crs_m <- 3310  # CA Albers coordinate system (meters)
grid_sf <- st_transform(grid_sf_ll, crs_m)
ca_bg  <- st_transform(ca_bg_shp, crs_m)

# rasterize
pts <- terra::vect(grid_sf)

r <- terra::rast(terra::ext(pts), res = 1000, crs = terra::crs(pts))  # 1km cells
pm_r <- terra::rasterize(pts, r, field = "pm", fun = mean, na.rm = TRUE)

# sanity checks (these should NOT look like 1 cell)
print(pm_r)
global(pm_r, "mean", na.rm = TRUE)

# 5) Polygon means
ca_bg_pm25 <- ca_bg[, "GEOID"]
ca_bg_pm25$pm25 <- exactextractr::exact_extract(pm_r, ca_bg_pm25, "mean")

summary(ca_bg_pm25$pm25)

# Plot as image 
png("../figs/ca_bg_pm25.png", width = 6, height = 5, units = "in", res = 300)
ggplot() +
  geom_sf(data = ca_bg_pm25, aes(fill = pm25), color = NA) +
  scale_fill_viridis_c(name = "PM2.5 (µg/m³)", na.value = "lightgrey") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_line(color = "transparent"),
    legend.background = element_rect(fill = alpha("white", 0.6))
  )
dev.off()

```

## Extract ACS 5 Year Census Data from 2012-2016
```{r}
# Set Census API key before proceeding
# Variable names
vars <- c(
  # income / poverty
  med_hh_income = "B19013_001", # in last year's inflation-adjusted dollars
  med_home_value = "B25077_001", # 
  pov_den = "B17010_001", pov_num = "B17010_002", # POVERTY STATUS IN THE PAST 12 MONTHS
  # employment
  emp_den     = "B23025_001", emp_num    = "B23025_002", # EMPLOYMENT STATUS FOR THE POPULATION 16 YEARS AND OVER
  # education (25+)
  HS_diploma    = "B15003_017", HS_equivalent = "B15003_018",
  some_college  = "B15003_019", some_college_plus = "B15003_020",
  assoc        = "B15003_021",
  bach          = "B15003_022", master       = "B15003_023",
  prof          = "B15003_024", doct         = "B15003_025", edu_den = "B15003_001",
  # tenure
  owner_occ     = "B25003_002", tenure_den   = "B25003_001", # TENURE 
  # race / ethnicity
  black         = "B02001_003", race_den     = "B02001_001",   # Black alone / total
  hisp          = "B03003_003", hisp_den     = "B03003_001",   # Hispanic / total
  # total population
  total_pop     = "B01003_001", 
  # median year structure built
  med_year_built = "B25035_001",
  # age < 5
  age_denom = "B01001_001", age_u5_male = "B01001_003", age_u5_female = "B01001_027",
  # age > 65
  age_6566_male = "B01001_020", age_6769_male = "B01001_021", age_7074_male = "B01001_022",
  age_7579_male = "B01001_023", age_8084_male = "B01001_024", age_85up_male = "B01001_025",
  age_6566_female = "B01001_044", age_6769_female = "B01001_045", age_7074_female = "B01001_046",
  age_7579_female = "B01001_047", age_8084_female = "B01001_048", age_85up_female = "B01001_049"
)

expected_E_cols <- paste0(names(vars), "E")

pull_one_year <- function(y) {
  # Only request variables that are available
  lv <- load_variables(y, "acs5", cache = TRUE)
  present <- names(vars)[vars %in% lv$name]
  vars_y  <- vars[present]
  if (length(vars_y) == 0) return(tibble())
  
  df <- get_acs(
    geography = "block group",
    state = "CA", 
    variables = vars_y,
    year = y, survey = "acs5",
    cache_table = TRUE, geometry = TRUE, output = "wide"
  )
  
  # Back-fill any missing expected columns with NA 
  for (col in setdiff(expected_E_cols, names(df))) df[[col]] <- NA_real_
  
  df %>%
    transmute(
      GEOID, NAME, year = y,
      
      med_hh_income = med_hh_incomeE,
      med_home_value = med_home_valueE,
      pov_rate      = ifelse(pov_denE     > 0, pov_numE / pov_denE, NA_real_),
      emp_rate    = ifelse(emp_denE   > 0, emp_numE   / emp_denE,   NA_real_),
      
      grad_HS     = ifelse(edu_denE     > 0, (HS_diplomaE + HS_equivalentE + some_collegeE + some_college_plusE + assocE +
                                                bachE + masterE + profE + doctE) / edu_denE, NA_real_),
      
      homeown_rate  = ifelse(tenure_denE  > 0, owner_occE   / tenure_denE,  NA_real_),
      
      pct_black     = ifelse(race_denE    > 0, blackE       / race_denE,    NA_real_),
      pct_hispanic  = ifelse(hisp_denE    > 0, hispE        / hisp_denE,    NA_real_),
      total_pop = total_popE,
      med_year_built = med_year_builtE,
      pct_under5 = ifelse(age_denomE > 0, (age_u5_maleE + age_u5_femaleE) / age_denomE, NA_real_),
      pct_over65 = ifelse(age_denomE > 0, (age_6566_maleE + age_6769_maleE + age_7074_maleE +
                                             age_7579_maleE + age_8084_maleE + age_85up_maleE +
                                             age_6566_femaleE + age_6769_femaleE + age_7074_femaleE +
                                              age_7579_femaleE + age_8084_femaleE + age_85up_femaleE) / age_denomE, NA_real_)
      
    )
}
# Pull just 2016 california
ca_census_2016 <- pull_one_year(2016)
```

## Merge the census data with the PM2.5 data
```{r}
mean(ca_census_2016$GEOID %in% ca_bg_pm25$GEOID) # 1
mean(ca_bg_pm25$GEOID %in% ca_census_2016$GEOID) # 1
ca_pm25 <- merge(ca_bg_pm25, st_drop_geometry(ca_census_2016), by = "GEOID")
ca_pm25 <- ca_pm25 %>%
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    pov = pov_rate,
    density = total_pop / area_km2
  )
# Get centroids of each block group and save coordinates as well
ca_pm25 <- ca_pm25 %>%
  mutate(
    centroid = st_centroid(geometry),
    cent_x = st_coordinates(centroid)[,1],
    cent_y = st_coordinates(centroid)[,2]
  ) %>%
  select(-centroid)

summary(ca_pm25)


write.csv(st_drop_geometry(ca_pm25), "data/ca_bg_pm25_census_2016.csv", row.names = F)
```

## Print five rows at random
```{r}
# Sample 5 rows at random, PM2.5, poverty, density
set.seed(123)
print(ca_pm25 %>%
  st_drop_geometry %>%
  select(GEOID, pov, density, pm25) %>% 
  sample_n(5), row.names = F)
```

## Generate synthetic outcome
```{r}
set.seed(1234)
ca_pm25 <- read.csv("data/ca_bg_pm25_census_2016.csv")

mu <- 0.25*ca_pm25$pov + 
  0.05*ca_pm25$pm25 + 
  0.1*log(ca_pm25$density) -
  0.3*ca_pm25$pct_hispanic +
  rnorm(nrow(ca_pm25),0,1)
# Generate disease counts using population (ca_pm25$total_pop) and mu
ca_pm25$Y <- rpois(nrow(ca_pm25), exp(mu)/500 * ca_pm25$total_pop)
summary(ca_pm25$Y)

# Run glm poisson
fit_glm <- glm(Y ~ pm25 + pov + log(density) + pct_hispanic,
               offset = log(total_pop),
               family = "poisson",
               data = ca_pm25)
summary(fit_glm) # ORACLE: coefficients look good
```

## Step 1: Fit a
```{r}
library(mgcv)
f_X_hat <- gam(pm25 ~ s(cent_x, # x coordinate of block group centroid
                        cent_y, # y coordinate of block group centroid
                        k = 300, # basis dimension
                        fx = T), # apply smoothing
               data = ca_pm25,
               method = 'GCV.Cp') # Generalized Cross Validation
ca_pm25$smoothed_pm25 <- f_X_hat$fitted.values
ca_pm25$residuals_pm25 <- f_X_hat$residuals
summary(f_X_hat)
```

## Step 2
```{r}
fit_spatial_plus <- mgcv::gam(
  Y ~ residuals_pm25 + # residuals of PM 2.5 from the previous regression
    pov + log(density) + # covariates
    s(cent_x, cent_y, k = 300, fx = T), # another thin plate spline of coordinates
  family = poisson(link = "log"), # Outcome is disease rate
  offset = log(total_pop), # offset of population
  data = ca_pm25,
  method = "GCV.Cp"
)
```

```{r}
summary(fit_spatial_plus)
# Save coefficient of residuals_pm25 and confidence interval in a data frame
results <- data.frame(
  term = "residuals_pm25",
  estimate = coef(fit_spatial_plus)["residuals_pm25"],
  conf_low = confint(fit_spatial_plus)["residuals_pm25", 1],
  conf_high = confint(fit_spatial_plus)["residuals_pm25", 2]
)
```


```{r}
# Compare to ordinary glm
fit_ols <- glm(Y ~ pm25 + pov + log(density) + pct_hispanic,
               offset = log(total_pop),
               family = "poisson",
               data = ca_pm25)
```


## Plot PM2.5, the fitted values, and residuals
```{r}
g1 <- ggplot() +
  geom_sf(data = ca_pm25, aes(fill = pm25), color = NA, linewidth = NA) +
  scale_fill_viridis_c(name = "PM2.5 (µg/m³)", na.value = "lightgrey") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_line(color = "transparent"),
    legend.background = element_rect(fill = alpha("white", 0.6)),
    legend.position = "bottom"
  )
g2 <- ggplot() +
  geom_sf(data = ca_pm25, aes(fill = smoothed_pm25), color = NA, linewidth = NA) +
  scale_fill_viridis_c(name = "Smoothed PM2.5 (µg/m³)", na.value = "lightgrey") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_line(color = "transparent"),
    legend.background = element_rect(fill = alpha("white", 0.6)),
    legend.position = "bottom"
  )
g3 <- ggplot() +
  geom_sf(data = ca_pm25, aes(fill = residuals_pm25), color = NA, linewidth = NA) +
  scale_fill_viridis_c(name = "Residuals (µg/m³)", na.value = "lightgrey") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_line(color = "transparent"),
    legend.background = element_rect(fill = alpha("white", 0.6)),
    legend.position = "bottom"
  )
png("../figs/ca_bg_pm25_spatialplus.png", width = 12, height = 5, units = "in", res = 300)
grid.arrange(g1, g2, g3, ncol = 3)
dev.off()
```

